---
title: "0.1 Data quality check"
author: "Dr. Chih-Yu Hung"
date: "2025-07-24"
output: html_document
---

## Check the model applcapbility and quality of output


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source("setup.R")
US_Data <- read_csv(us_outputs("0.4_US_Cinput.csv"))

CA_Data <- read_rds("Input/Canada Cinput_5y.rds") %>%
  filter(CROPS %in% c("BARLEY", "CANOLA", "CORNGR",
                      "CORNSLG", "DFPEAS", "RYEFAL",
                      "WHTSPG", "WHTDUR", "WHTWINT", 
                      "SOYBNS", "TRICTCL", "RYESPG")) %>%
  mutate(UTIL_PRACTICE_DESC = case_when(
    CROPS ==  "CORNGR"      ~ "GRAIN",
    CROPS ==  "CORNSLG"     ~ "SILAGE",
    TRUE                    ~ "ALL UTILIZATION PRACTICES"
  )) %>%
  mutate(CLASS = case_when(
    CROPS ==  "WHTSPG"      ~ "SPRING, (EXCL DURUM)",
    CROPS ==  "WHTWINT"     ~ "WINTER",
    CROPS ==  "WHTDUR"      ~ "SPRING, DURUM",
    CROPS ==  "DFPEAS"      ~ "DRY EDIBLE",
    TRUE                    ~ "ALL CLASEES"
  )) %>%
  mutate(CROPS = case_when(
    CROPS %in%  c("CORNGR","CORNSLG")           ~ "CORN",
    CROPS %in%  c("WHTSPG","WHTWINT","WHTDUR")  ~ "WHEAT",
    CROPS %in%  c("RYEFAL", "RYESPG")           ~ "RYE",
    CROPS ==  "DFPEAS"                          ~ "PEAS",
    CROPS ==  "TRICTCL"                         ~ "TRITICALE",
    CROPS ==  "SOYBNS"                          ~ "SOYBEANS",
    TRUE                                        ~  CROPS
  )) %>%
  select(STATE = Province_ID, COUNTY = ECODISTRIC, CROP = CROPS, CLASS, UTIL_PRACTICE_DESC,
         AREA, Cinput, MAP, MAT, MPE, CARB30, Texture = TEXCAT, Frac)

SiteData <- rbind (US_Data, CA_Data) %>%
    mutate(TEXT = case_when(
    Texture %in% c("fine","Fine")        ~ 3,
    Texture %in% c("medium","Medium")    ~ 2,
    Texture %in% c("coarse","Coarse")    ~ 1,
    TRUE                   ~NA_integer_
  ))

```

## Check the % of production within Canada's MAP, MAT...

### Compare the US and CA data together

64.5% of area in the similar environment. The production could be higher if we consider the large amount of corn in CA.

If we go to MAT 14 degree C, We include 84% of production, 15 degree C, 88% of production.

### If compare the data in the US only (Code not show here)
67% and 61% of crop production and area in similar environment, respectively.

If we go to MAP 14 degree C, We include 86% of production, 15 degree C, 90% of production.

```{r }

Area <- SiteData %>% 
      filter(MAT >= -1 & MAT <= 11) %>%
      filter(MAP >= 300 & MAP <= 2500)


sum(Area$AREA) / sum(SiteData$AREA) # 64.5% of total area
```

## Run the RF model


```{r }
load("Output/Model_5.RData") # ML model from 5-y moving average.

SiteData$SOC_predict <- predict(rf_model, newdata = SiteData)

#CAP deltaSOC to Cinput
SiteData <- SiteData %>%
  mutate(SOC_predict = ifelse(SOC_predict > Cinput, Cinput, SOC_predict))


#group the data to STATE/Province level
Result <- SiteData %>%
  mutate(CROP_ID = case_when(
    CROP == "CORN"          ~ paste0(CROP, "_", UTIL_PRACTICE_DESC, ""),
    CROP == "WHEAT"         ~ paste0(CROP, "_", CLASS, ""),
    TRUE                    ~ CROP 
  ))


#helper for weighted means that gracefully handles all-NA cases
w_mean <- function(x, w) {
  if (all(is.na(x))) return(NA_real_)
  weighted.mean(x, w, na.rm = TRUE)
}

#obtain weighted values and simple average
Result_State <- Result %>% 
  mutate(area_w = AREA * Frac / 100) %>%          # weight column once
  group_by(STATE, CROP_ID) %>% 
  summarise(
    across(
      c(Cinput, MAT, MAP, MPE, CARB30, SOC_predict),
      list(avg  = ~mean(.x, na.rm = TRUE),
           wavg = ~w_mean(.x, area_w)),
      .names = "{.col}_{.fn}"
    ),
    .groups = "drop"                              # remove grouping (optional)
  )%>% 
  select(STATE, CROP_ID, Cinput_wavg, MAT_wavg, MAP_wavg, MPE_wavg, CARB30_wavg,SOC_predict_wavg)

```


## draw maps to see the SOC change across North America

```{r}
library(tidyverse)
library(sf)
library(rnaturalearth)
library(rnaturalearthhires)

# ───────────────────────── 1. polygons ─────────────────────────────
na_states <- ne_states(returnclass = "sf") %>%        # provinces + states
  filter(admin %in% c("United States of America", "Canada")) %>%
  select(name = name_en, postal, geometry)

# ───────────────────────── 2. tidy keys ────────────────────────────
us_lookup <- tibble(
  STATE      = toupper(state.name),      # <- uppercase!
  postal     = state.abb
)

Result_State_prepped <- Result_State %>%
  mutate(STATE = toupper(STATE)) %>%      # <- uppercase!
  left_join(us_lookup, by = "STATE") %>%
  mutate(postal = coalesce(postal, STATE)) %>%   # US abb or existing CA code
  select(postal, CROP_ID, SOC_predict_wavg)      # keep what we need

# ──────────────────────── 3. one row per region-crop ───────────────
Result_State_single <- Result_State_prepped %>%
  group_by(postal, CROP_ID) %>%
  summarise(SOC_predict_wavg = mean(SOC_predict_wavg, na.rm = TRUE),
            .groups = "drop")

plot_data <- na_states %>%
  left_join(Result_State_single, by = "postal") %>%
  na.omit()


# all 13 crop IDs present in your dataset ────────────────────────────────
all_crops <- unique(plot_data$CROP_ID)

# ── 2. distinct polygons (one row per province / state) ───────────────────────
polygons_only <- plot_data %>%                      # keep name, postal & geometry
  select(name, postal, geometry) %>% 
  distinct()

# ── 3. Cartesian product  (every polygon × every crop) ───────────────────────
plot_data_full <- polygons_only %>%                 # start with polygons
   crossing(CROP_ID = all_crops) %>%                # replicate 13×
   left_join(plot_data, by = c("name", "postal", "CROP_ID", "geometry")
  )

# ── make the crossed table an sf object again ─────────────────────────
plot_data_full <- st_as_sf(
  plot_data_full,                     # your tibble with a ‘geometry’ column
  crs = st_crs(polygons_only)         # re-attach the CRS from the source map
)


# ───────────────────────── 4. plot ─────────────────────────────────
ggplot(plot_data_full) +
  geom_sf(aes(fill = SOC_predict_wavg), colour = "grey70", size = 0.1) +
  facet_wrap(~ CROP_ID) +
  scale_fill_gradient2(
    name       = "delta SOC (t/ha)",
    low        = "#00008B",   # deep blue for SOC losses
    mid        = "#ADD8E6",   # light-blue for ~0
    high       = "#FFFF00",   # blue for SOC gains
    midpoint   = 0,
    na.value   = "white"      # NA polygons → white
  ) +
  coord_sf(crs = 3347) +       # Canada Albers equal-area (good for all NA)
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.major = element_blank(),
    strip.text       = element_text(face = "bold"),
    legend.position  = "right"
  )


ggsave("Figures/SOC_maps_by_crop.png", width = 10, height = 8)


```
