---
title: "Delta SOC_Crop input"
author: "Dr. Chih-Yu Hung"
date: "2024-06-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(polycor)
library(mgcv)
library(caret)
library(zoo)
library(gridExtra)
```


## Build model 

### Linear model, Stepwise

```{r stepwise model}
#full_model <- lm(deltaSOC ~ .-Year -SiteName -Province - Ecodistrict - Area, data = SiteData)
#stepwise_model <- step(full_model, direction = "backward")
#summary(stepwise_model)
#cat("R² value: ", summary(stepwise_model)$r.squared, "\n")

load("Input/SiteData_Daycent_5y.RData")

SiteData <- SiteData_5y


model <- lm(deltaSOC ~ MAP * CARB30 * Cinput * TEXT * Region , data = SiteData)
stepwise_model <- step(model, direction = "backward")
summary(stepwise_model)

```


### Generalized additive model

```{r Generalized additive model}
# Fit a GAM model
#model_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(TEXT, k=5) + s(CARB30) + s(Region, k = 5), data = SiteData)
#model_gam_grow <- gam(deltaSOC ~ s(MAP) + s(MAT_grow) + s(MPE) + s(Cinput) + s(TCLAY) + s(TSAND) + s(CARB30), data = SiteData)

# Summary of the model
#summary(model_gam) #R2 = 27.4
#summary(model_gam_grow) #R2 = 27.7

```


```{r model comparision}
# AIC comparison
#AIC(model, model_gam) #barely different
#AIC(model_gam,model_gam_grow)#barely different
```


## Group data and redo the regression models

In this part, we will group the data into three soil texture groups (Fine, Medium, Coarse),


### Soil textured group-prepare train-test split

```{r Grouped model- Soil texture}
#1. Soil texture group

SiteData_Fine <- SiteData %>%
  filter(TEXT == 3)
SiteData_Medium <- SiteData %>%
  filter(TEXT == 2)
SiteData_Coarse <- SiteData %>%
  filter(TEXT == 1)


#Set the train-test split
set.seed(2024)

trainId_Fine <- createDataPartition(SiteData_Fine$deltaSOC, p = 0.8, list = FALSE)
trainId_Medium <- createDataPartition(SiteData_Medium$deltaSOC, p = 0.8, list = FALSE)
trainId_Coarse <- createDataPartition(SiteData_Coarse$deltaSOC, p = 0.8, list = FALSE)

trainFine <- SiteData_Fine[trainId_Fine,]
trainMedium <- SiteData_Medium[trainId_Medium,]
trainCoarse <- SiteData_Coarse[trainId_Coarse,]

testFine <- SiteData_Fine[-trainId_Fine,]
testMedium <- SiteData_Medium[-trainId_Medium,]
testCoarse <- SiteData_Coarse[-trainId_Coarse,]

```


### Soil textured group- regression for the training group

```{r regression model for soil texture groups}
## Going to use GAM model only
# #Fit the regression models
# model_Fine <- lm(deltaSOC ~ MAP + MAT + CARB30 + Cinput + Region , data = SiteData_Fine)
# #model_Fine <- lm(deltaSOC ~ MAP * MAT * CARB30 * Cinput * Region , data = SiteData_Fine)
# stepwise_Fine <- step(model_Fine, direction = "backward")
# summary(stepwise_Fine) # R2 = 0.451
# 
# model_Med <- lm(deltaSOC ~ MAP + MAT + CARB30 + Cinput + Region , data = SiteData_Medium)
# model_Med <- lm(deltaSOC ~ MAP * MAT * CARB30 * Cinput * Region , data = SiteData_Medium)
# stepwise_Med <- step(model_Med, direction = "backward")
# summary(stepwise_Med) #R2 = 0.545
# 
# model_Coarse <- lm(deltaSOC ~ MAP + MAT + CARB30 + Cinput + Region , data = SiteData_Coarse)
# model_Coarse <- lm(deltaSOC ~ MAP * MAT * CARB30 * Cinput * Region , data = SiteData_Coarse)
# stepwise_Coarse <- step(model_Coarse, direction = "backward")
# summary(stepwise_Coarse) #highest R2= 0.737


#GAM for training set
model_Fine_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(CARB30), data = trainFine)
#model_Fine_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput), data = trainFine) R2 =0.41
summary(model_Fine_gam) #R2 = 0.668

model_Medium_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(CARB30), data = trainMedium)
#model_Medium_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput), data = trainMedium) R2 = 0.43
summary(model_Medium_gam) #R2 = 0.658

model_Coarse_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(CARB30) , data = trainCoarse)
#model_Coarse_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) , data = trainCoarse) R2 =0.36
summary(model_Coarse_gam) #R2 = 0.782

#Save the GAM models
# save(model_Fine_gam,file = "Output/GAM model_Fine.RData")
# save(model_Medium_gam,file = "Output/GAM model_Medium.RData")
# save(model_Coarse_gam,file = "Output/GAM model_Coarse.RData")

#predict delta SOC with train data
predict_train_Fine <- predict(model_Fine_gam, newdata = trainFine)
predict_train_Medium <- predict(model_Medium_gam, newdata = trainMedium)
predict_train_Coarse <- predict(model_Coarse_gam, newdata = trainCoarse)


#predict delta SOC with test data
predict_test_Fine <- predict(model_Fine_gam, newdata = testFine)
predict_test_Medium <- predict(model_Medium_gam, newdata = testMedium)
predict_test_Coarse <- predict(model_Coarse_gam, newdata = testCoarse)


#Compare the predictions and actual values by using Mean square error (MSE)
MSE_Fine <- mean((predict_test_Fine-testFine$deltaSOC)^2)        # 0.0057 t/ha
MSE_Medium <- mean((predict_test_Medium-testMedium$deltaSOC)^2)  # 0.0040 t/ha
MSE_Coarse <- mean((predict_test_Coarse-testCoarse$deltaSOC)^2)  # 0.0018 t/ha

#Compare the predictions and actual values by using R2
R_squared <- function(actuals, predictions) {
  ss_res <- sum((actuals - predictions)^2)
  ss_tot <- sum((actuals - mean(actuals))^2)
  r_squared <- 1 - (ss_res / ss_tot)
  return(r_squared)
}

R2_Fine <- R_squared(predict_test_Fine, testFine$deltaSOC) #0.47
R2_Medium <- R_squared(predict_test_Medium, testMedium$deltaSOC) #0.48
R2_Coarse <- R_squared(predict_test_Coarse, testCoarse$deltaSOC) #0.76


#Compare the predictions and actual values by mean absolute percentage error (MAPE)
MAPE_Fine <- mean(abs(testFine$deltaSOC-predict_test_Fine)/testFine$deltaSOC)*100            #24.7%
MAPE_Medium <- mean(abs(testMedium$deltaSOC-predict_test_Medium)/testMedium$deltaSOC)*100    #9.1%
MAPE_Coarse <- mean(abs(testCoarse$deltaSOC-predict_test_Coarse)/testCoarse$deltaSOC)*100    #29.7%

```


### Plot the preidictions and actual data

```{r Plot the data, fig.height = 9}
#Train data
plot_train_Fine <- data.frame(actual = trainFine$deltaSOC, predictions = predict_train_Fine)
plot_train_Medium <- data.frame(actual = trainMedium$deltaSOC, predictions = predict_train_Medium)
plot_train_Coarse <- data.frame(actual = trainCoarse$deltaSOC, predictions = predict_train_Coarse)


#Test data
plot_test_Fine <- data.frame(actual = testFine$deltaSOC, predictions = predict_test_Fine)
plot_test_Medium <- data.frame(actual = testMedium$deltaSOC, predictions = predict_test_Medium)
plot_test_Coarse <- data.frame(actual = testCoarse$deltaSOC, predictions = predict_test_Coarse)



plot_comp <- function(plot_test_data, plot_train_data) {
  # Calculate MSE and MAPE for test data
  mse <- mean((plot_test_data$predictions - plot_test_data$actual)^2)
  mape <- mean(abs(plot_test_data$actual - plot_test_data$predictions) / plot_test_data$actual) * 100 
  dataset_name <- deparse(substitute(plot_test_data))
  
  x_max <- max(plot_train_data$actual)
  y_min <- min(plot_train_data$predictions)
  Soil_name <- sub(".*_", "", deparse(substitute(plot_test_data)))
    ggplot() +
    geom_point(data = plot_train_data, aes(x = actual, y = predictions), color = 'blue', shape = 1, show.legend = TRUE ) + # Train data in red
    geom_point(data = plot_test_data, aes(x = actual, y = predictions), color = 'red', shape = 1, show.legend = TRUE ) +  # Test data in blue
    geom_abline(slope = 1, intercept = 0, color = 'red', linetype = 'dashed') +  # 1:1 line
    labs(title = paste('Model vs GAM Values for', Soil_name),  # Add title and labels
         x = 'Model Values (tC /ha)',
         y = 'GAM Values (tC /ha)',
         color = "Data Type") +
    theme_minimal() +
    coord_cartesian(xlim = c(-0.7, 0.7), ylim = c(-0.7, 0.7))  # Set x and y axis limits
    #scale_color_manual(values = c("Test Data" = "red", "Train Data" = "blue")) #+
    # annotate("text", x = x_max, y = y_min*1.1, 
    #          label = paste("MSE:", round(mse, 4), "t C/ha"), 
    #          hjust = 1.1, vjust = 2, size = 4, color = "black") +  # Add MSE annotation
    # annotate("text", x = x_max, y = y_min * 1.2, 
    #          label = paste("MAPE:", round(mape, 2), "%"), 
    #          hjust = 1.1, vjust = 3.5, size = 4, color = "black")   # Add MAPE annotation
}

plot_F <- plot_comp(plot_test_Fine,plot_train_Fine)
plot_M <- plot_comp(plot_test_Medium,plot_train_Medium)
plot_C <- plot_comp(plot_test_Coarse,plot_train_Coarse)

grid.arrange(plot_F,plot_M,plot_C, heights = unit(c(3,3,3), "inches"))


```

### Model performance of training and test data

```{r model performance}
train.data <-list(plot_train_Fine, plot_train_Medium, plot_train_Coarse)
test.data <- list(plot_test_Fine, plot_test_Medium, plot_test_Coarse)

model.p <- function (data) {
  # Calculate MSE and MAPE for test data
  mse <- round(mean((data$predictions - data$actual)^2),3)
  mape <- round(mean(abs(data$actual - data$predictions) / data$actual) * 100,1) 
  return(list(MSE = mse, MAPE = mape))
}

# Apply the function to each dataset in train.data and test.data
train_results <- lapply(train.data, model.p)
test_results <- lapply(test.data, model.p)

# Combine the results into a data frame
results <- data.frame(
  Dataset = c("Fine", "Medium", "Coarse"),
  Train_MSE = sapply(train_results, function(x) x$MSE),
  Train_MAPE = sapply(train_results, function(x) x$MAPE),
  Test_MSE = sapply(test_results, function(x) x$MSE),
  Test_MAPE = sapply(test_results, function(x) x$MAPE)
)


knitr::kable(results, caption = "MSE and sMAPE for Train and Test Data")

```



### C input group (Obsolete).

The C input ranges were determined by Q1, Q2, Q3, and Q4


``` {r Grouped model - C input}
#2. C input group
# summary(SiteData$Cinput)
# SiteData_HighC <- SiteData %>%
#   filter(Cinput >= 3.11 )
# SiteData_MHC <- SiteData %>% 
#   filter(Cinput >= 2.538 & Cinput < 3.11 ) #Medium high
# SiteData_MLC <- SiteData %>%
#   filter(Cinput <= 2.538 & Cinput > 2.1  ) #Medium low
# SiteData_LowC <- SiteData %>%
#   filter(Cinput <= 2.1 )
# 
# SiteData_Clow <- SiteData %>%
#   filter( CARB30 > 6250  ) #Medium low
# SiteData_Chigh <- SiteData %>%
#   filter(CARB30 <= 6250 )



#Fit the regression models
# model_HighC <- lm(deltaSOC ~ MAP + MAT + CARB30 + Cinput + Region , data = SiteData_HighC)
# stepwise_HighC <- step(model_HighC, direction = "backward")
# summary(stepwise_HighC) #R2 = 0.554
# 
# model_MHC <- lm(deltaSOC ~ MAP + MAT + CARB30 + Cinput + Region , data = SiteData_MHC)
# stepwise_MHC <- step(model_MHC, direction = "backward")
# summary(stepwise_MHC) #R2 = 0.410
# 
# model_MLC <- lm(deltaSOC ~ MAP + MAT + CARB30 + Cinput + Region , data = SiteData_MLC)
# stepwise_MLC <- step(model_MLC, direction = "backward")
# summary(stepwise_MLC) #R2 = 0.392
# 
# model_LowC <- lm(deltaSOC ~ MAP + MAT + CARB30 + Cinput + Region , data = SiteData_LowC)
# stepwise_LowC <- step(model_LowC, direction = "backward")
# summary(stepwise_LowC) # R2 = 0.421

#GAM 
# model_HighC_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(CARB30), data = SiteData_HighC)
# summary(model_HighC_gam) #R2 = 0.700
# 
# model_MHC_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(CARB30), data = SiteData_MHC)
# summary(model_MHC_gam) #R2 = 0.615
# 
# model_MLC_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(CARB30), data = SiteData_MLC)
# summary(model_MLC_gam) #R2 = 0.617
# 
# model_LowC_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(CARB30), data = SiteData_LowC)
# summary(model_LowC_gam) #R2 = 0.598

```

### MAP group (Obsolete)

The MAPs were determined arbitary


```{r Group model - MAP}
#3. MAP group
# SiteData_HighP <- SiteData %>%
#   filter(MAP > 1000 )
# SiteData_MedP <- SiteData %>%
#   filter(MAP > 500 & MAP <= 1000 )
# SiteData_LowP <- SiteData %>%
#   filter(MAP <= 500 )

#Fit the regression models
# model_HighP <- lm(deltaSOC ~ MAP + MAT + CARB30 + Cinput + Region , data = SiteData_HighP)
# stepwise_HighP <- step(model_HighP, direction = "backward")
# summary(stepwise_HighP) #R2 0.534
# 
# model_MedP <- lm(deltaSOC ~ MAP + MAT + CARB30 + Cinput + Region , data = SiteData_MedP)
# stepwise_MedP <- step(model_MedP, direction = "backward")
# summary(stepwise_MedP) #R2 0.543
# 
# model_LowP <- lm(deltaSOC ~ MAP + MAT + CARB30 + Cinput + Region , data = SiteData_LowP)
# stepwise_LowP <- step(model_LowP, direction = "backward")
# summary(stepwise_LowP) #R2 = 0.529


#GAM 
# model_HighP_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(CARB30), data = SiteData_HighP)
# summary(model_HighP_gam) #R2 = 0.692
# 
# model_MedP_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(CARB30), data = SiteData_MedP)
# summary(model_MedP_gam) #R2 = 0.684
# 
# model_LowP_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(CARB30), data = SiteData_LowP)
# summary(model_LowP_gam) #R2 = 0.678

```


## Make a Lookup table (Obsolete)
## Abandoned for moving average analysis
The look-up table include MAP, MAT, Soil Texture Class and C input
Based on the regression results, I believe =


```{r Lookup table }
# Ensure TEXT is a factor
# SiteData$TEXT <- as.factor(SiteData$TEXT)
# SiteData<-na.omit(SiteData)

# Separate to Eastern and Western Canada
# SiteData_West <- SiteData[SiteData$Region == 1, ]
# SiteData_East <- SiteData[SiteData$Region == 2, ]

#Bin for West part
# Bin MAP
# SiteData_West$MAP_bin <- cut(SiteData_West$MAP, breaks = 2, labels = FALSE)
# Bin MAT
# SiteData_West$MAT_bin <- cut(SiteData_West$MAT, breaks = 2, labels = FALSE)
# Bin Cinput
# SiteData_West$Cinput_bin <- cut(SiteData_West$Cinput, breaks = 2, labels = FALSE)
# Bin initial C
# SiteData_West$CARB30_bin <- cut(SiteData_West$CARB30, breaks = 2, labels = FALSE)

# Create lookup table_West
# lookuptable_West <- SiteData_West %>%
#   group_by(TEXT, MAP_bin, MAT_bin, Cinput_bin, CARB30_bin) %>%
#   summarize(mean_deltaSOC = mean(deltaSOC, na.rm = TRUE),
#             median_deltaSOC = median(deltaSOC, na.rm = TRUE),
#             sd_deltaSOC = sd(deltaSOC, na.rm = TRUE),
#             n = n()) %>%
#   ungroup()

#Bin for East part
# Bin MAP
# SiteData_East$MAP_bin <- cut(SiteData_East$MAP, breaks = 2, labels = FALSE)
# Bin MAT
# SiteData_East$MAT_bin <- cut(SiteData_East$MAT, breaks = 2, labels = FALSE)
# Bin Cinput
# SiteData_East$Cinput_bin <- cut(SiteData_East$Cinput, breaks = 2, labels = FALSE)
# Bin initial C
# SiteData_East$CARB30_bin <- cut(SiteData_East$CARB30, breaks = 2, labels = FALSE)

# Create lookup table_East
# lookuptable_East <- SiteData_East %>%
  # group_by(TEXT, MAP_bin, MAT_bin, Cinput_bin, CARB30_bin) %>%
  # summarize(mean_deltaSOC = mean(deltaSOC, na.rm = TRUE),
  #           median_deltaSOC = median(deltaSOC, na.rm = TRUE),
  #           sd_deltaSOC = sd(deltaSOC, na.rm = TRUE),
  #           n = n()) %>%
  # ungroup()


#Write the csv files
#write.csv(lookuptable_West,"output/lookuptable_West.csv")
#write.csv(lookuptable_East,"output/lookuptable_East.csv")


```

