---
title: "Daycent C Data preparation"
author: "Dr. Chih-Yu Hung"
date: "2024-07-04"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(polycor)
library(mgcv)
library(caret)
library(zoo)
library(gridExtra)
```

## Daycent data preparation

Canada’s NIR have estimate annual ΔSOC at ecodistrict level. According to Thiagarajan et al., (2022), ΔSOC have a strong relationship with site specific variables and C inputs. We therefore propose to build a regression model to predict ΔSOC. The regression model could have variables including mean annual precipitation, mean air temperature, mean temperature in growing seasons, C input, initial C stock, and soil texture. 

This document organized the data from AAFC (ElKhoury, Antoun (AAFC/AAC) <antoun.elkhoury@AGR.GC.CA>).
Files include:

1) CinputsAtEcodprov_19Oct2023: the ecodistrict C input used for the model run.
2) ecodcinputbycrop:            the cinput values by crop after adjustment for residue removal.
3) EcoProvTexSites:             the carbon initialization values (note carbon data is in g/m2).
4) SiteAdjDelta_19Oct2023:      the inventory delta C output from 1990 to 2022.
5) slyieldadj19102023:          the adjusted yield data used for the model run.
6) weather_master:              the climate file.

```{r Loading data}
Cinput    <- read.csv("Input/CropC_Daycent/CinputsAtEcodprov_19Oct2023.csv") #C input proportion not crop specific
Ecodc     <- read.csv("Input/CropC_Daycent/ecodcinputbycrop.csv") # C input crop specific 
EcoP      <- read.csv("Input/CropC_Daycent/EcoProvTexSites.csv") # C ini values
SiteAdj   <- read.csv("Input/CropC_Daycent/SiteAdjDelta_19Oct2023.csv") # Delta C not crop specific
Slyield   <- read.csv("Input/CropC_Daycent/slyieldadj19102023.csv") # yield data, no need
Weather_y <- read.csv("Input/CropC_Daycent/weather_master.csv") #weahter

#prepare weather data from monthly to yearly, the table will have
#Ecodistrict, MAP, MAT, MAP_grow, MAP_grow, MPE(mean potential evapotranspiration) and MPE_grow. The _grow means average values in May-Oct. 

grow <- c(5:10) #Define growing season in May-Oct
Weather <- Weather_y %>%
  select(Ecodistrict = ECODISTRIC, month, pe, precip, temp) %>%
  group_by(Ecodistrict) %>%
  summarise(MAP = sum(precip),MAT = mean(temp),MPE = sum(pe),
            MAP_grow = sum(precip[month %in% grow]),
            MAT_grow = mean(temp[month %in% grow]),
            MPE_grow = sum(pe[month %in% grow]))
#Save data
save(Weather, file = "Input/Data_Weather.RData")                           

#Prepare C input based on C input proportion and crop specific C input(t)
TotalCinput <- Ecodc %>%
  group_by(ECODISTRIC, YEAR, Province_ID) %>%
  summarise(Cinput = sum(CInput_withRootExudates_t)/sum(Corrected_Area))
```


## Making Dataframe, prepare for model_Use the Avg. 5 y data only

### Major file: `SiteAdj`, the `deltaSOCAdj` (t/ha, not crop specific) is the final delta SOC

```{r Data preparation}

SiteData <- SiteAdj %>%
  select(Year = year, SiteName, Province = Province_ID, Ecodistrict = ECODISTRIC, Area = Area_per_site_Ha, deltaSOC = deltaSOCAdj )

#join the weather data
SiteData <- SiteData %>%
  inner_join(Weather, by = c("Ecodistrict" = "Ecodistrict" )) %>%
  inner_join(EcoP, by = c("Ecodistrict" =  "ECODISTRIC", "Province" = "PROVINCE","SiteName" = "sitename"))
  

#join the Cinput
SiteData <- SiteData %>%
  inner_join(TotalCinput, by = c( "Ecodistrict" = "ECODISTRIC", "Province" = "Province_ID", "Year" = "YEAR"))


#NA omit, there are some data has no soil texture categories. (TEXCAT = na)
SiteData <- na.omit(SiteData)
SiteData <- SiteData[!(SiteData$TEXCAT == "na"),]

#Put dummy variables for soil texture and Eastern and Western Canada
#We believe crops in Western and Eastern Canada are quite different and have different C/N that could influence delta SOC
SiteData <- SiteData %>%
mutate(TEXT = case_when(
TEXCAT == "coarse" ~ 1,
TEXCAT == "medium" ~ 2,
TEXCAT == "fine" ~ 3,
TRUE ~ 99)) %>%
mutate(Region = case_when(
Province %in% c("BC","AB","SK","MB") ~ 1,
TRUE ~ 2))

#Save data
#save(SiteData, file="Input/SiteData_Daycent.RData")

#Take the avg. for the moving avg. 5 yr. 
SiteData_5y <- SiteData %>%
  group_by(SiteName) %>%
  select(SiteName,deltaSOC, MAP, MAT, MPE, CARB30, Cinput, TEXT, Region) %>%
  mutate(deltaSOC = rollmean(deltaSOC, k = 5, fill= NA, align = "right")) %>%
  mutate(Cinput = rollmean(Cinput, k = 5, fill= NA, align = "right")) %>%
  na.omit()

#Save data
#save(SiteData_5y, file="Input/SiteData_Daycent_5y.RData")
```



### prepare the data for GAMs

I need the avg. data for the most recent 5 y

```{r data preparation for GMAs}
#Fine texture data
Data_GAM_Fine <- SiteData %>%
  filter(Year >= 2018 & TEXT == 3) %>%
  group_by(SiteName) %>%
  select(Province, SiteName,deltaSOC, MAP, MAT, MPE, CARB30, Cinput) %>%
  mutate(deltaSOC = rollmean(deltaSOC, k = 5, fill= NA, align = "right")) %>%
  mutate(Cinput = rollmean(Cinput, k = 5, fill= NA, align = "right")) %>%
  na.omit()

#Medium texture data
Data_GAM_Medium <- SiteData %>%
  filter(Year >= 2018 & TEXT == 2) %>%
  group_by(SiteName) %>%
  select(Province, SiteName,deltaSOC, MAP, MAT, MPE, CARB30, Cinput) %>%
  mutate(deltaSOC = rollmean(deltaSOC, k = 5, fill= NA, align = "right")) %>%
  mutate(Cinput = rollmean(Cinput, k = 5, fill= NA, align = "right")) %>%
  na.omit()

#Coarse texture data
Data_GAM_Coarse <- SiteData %>%
  filter(Year >= 2018 & TEXT == 1) %>%
  group_by(SiteName) %>%
  select(Province, SiteName, deltaSOC, MAP, MAT, MPE, CARB30, Cinput) %>%
  mutate(deltaSOC = rollmean(deltaSOC, k = 5, fill= NA, align = "right")) %>%
  mutate(Cinput = rollmean(Cinput, k = 5, fill= NA, align = "right")) %>%
  na.omit()

#Save the data 
save(Data_GAM_Fine, file = "Input/Data_GAM_Fine.RData")
save(Data_GAM_Medium, file = "Input/Data_GAM_Medium.RData")
save(Data_GAM_Coarse, file = "Input/Data_GAM_Coarse.RData")

```


### prepare the data for GAMs only 2017-2019 to check CRC data

```{r data preparation for GMAs}
#Fine texture data
Data_GAM_Fine_CRC <- SiteData %>%
  filter(Year >= 2017 & Year <=2019 & TEXT == 3) %>%
  group_by(SiteName) %>%
  select(Province, SiteName,deltaSOC, MAP, MAT, MPE, CARB30, Cinput) %>%
  mutate(deltaSOC = rollmean(deltaSOC, k = 3, fill= NA, align = "right")) %>%
  mutate(Cinput = rollmean(Cinput, k = 3, fill= NA, align = "right")) %>%
  na.omit()

#Medium texture data
Data_GAM_Medium_CRC <- SiteData %>%
  filter(Year >= 2017 & Year <=2019 & TEXT == 2) %>%
  group_by(SiteName) %>%
  select(Province, SiteName,deltaSOC, MAP, MAT, MPE, CARB30, Cinput) %>%
  mutate(deltaSOC = rollmean(deltaSOC, k = 3, fill= NA, align = "right")) %>%
  mutate(Cinput = rollmean(Cinput, k = 3, fill= NA, align = "right")) %>%
  na.omit()

#Coarse texture data
Data_GAM_Coarse_CRC <- SiteData %>%
  filter(Year >= 2017 & Year <=2019 & TEXT == 1) %>%
  group_by(SiteName) %>%
  select(Province, SiteName, deltaSOC, MAP, MAT, MPE, CARB30, Cinput) %>%
  mutate(deltaSOC = rollmean(deltaSOC, k = 3, fill= NA, align = "right")) %>%
  mutate(Cinput = rollmean(Cinput, k = 3, fill= NA, align = "right")) %>%
  na.omit()

#Save the data 
save(Data_GAM_Fine_CRC, file = "Input/Data_GAM_Fine_2017-2019.RData")
save(Data_GAM_Medium_CRC, file = "Input/Data_GAM_Medium_2017-2019.RData")
save(Data_GAM_Coarse_CRC, file = "Input/Data_GAM_Coarse_2017-2019.RData")

```