---
title: "Delta SOC_Crop input"
author: "Dr. Chih-Yu Hung"
date: "2024-05-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(polycor)
library(mgcv)
library(caret)

```

## Daycent data preparation

Canada’s NIR have estimate annual ΔSOC at ecodistrict level. According to Thiagarajan et al., (2022), ΔSOC have a strong relationship with site specific variables and C inputs. We therefore propose to build a regression model to predict ΔSOC. The regression model could have variables including mean annual precipitation, mean air temperature, mean temperature in growing seasons, C input, initial C stock, and soil texture. 

This document organized the data from AAFC (ElKhoury, Antoun (AAFC/AAC) <antoun.elkhoury@AGR.GC.CA>).
Files include:

1) CinputsAtEcodprov_19Oct2023: the ecodistrict C input used for the model run.
2) ecodcinputbycrop:            the cinput values by crop after adjustment for residue removal.
3) EcoProvTexSites:             the carbon initialization values (note carbon data is in g/m2).
4) SiteAdjDelta_19Oct2023:      the inventory delta C output from 1990 to 2022.
5) slyieldadj19102023:          the adjusted yield data used for the model run.
6) weather_master:              the climate file.

```{r Loading data}
Cinput    <- read.csv("Input/CropC_Daycent/CinputsAtEcodprov_19Oct2023.csv") #C input proportion not crop specific
Ecodc     <- read.csv("Input/CropC_Daycent/ecodcinputbycrop.csv") # C input crop specific 
EcoP      <- read.csv("Input/CropC_Daycent/EcoProvTexSites.csv") # C ini values
SiteAdj   <- read.csv("Input/CropC_Daycent/SiteAdjDelta_19Oct2023.csv") # Delta C not crop specific
Slyield   <- read.csv("Input/CropC_Daycent/slyieldadj19102023.csv") # yield data, no need
Weather_y <- read.csv("Input/CropC_Daycent/weather_master.csv") #weahter

#prepare weather data from monthly to yearly, the table will have
#Ecodistrict, MAP, MAT, MAP_grow, MAP_grow, MPE(mean potential evapotranspiration) and MPE_grow. The _grow means average values in May-Oct. 

grow <- c(5:10) #Define growing season in May-Oct
Weather <- Weather_y %>%
  select(Ecodistrict = ECODISTRIC, month, pe, precip, temp) %>%
  group_by(Ecodistrict) %>%
  summarise(MAP = sum(precip),MAT = mean(temp),MPE = sum(pe),
            MAP_grow = sum(precip[month %in% grow]),
            MAT_grow = mean(temp[month %in% grow]),
            MPE_grow = sum(pe[month %in% grow]))
                           
#Prepare C input based on C input proportion and crop specific C input(t)
TotalCinput <- Ecodc %>%
  group_by(ECODISTRIC, YEAR, Province_ID) %>%
  summarise(Cinput = sum(CInput_withRootExudates_t))
```


## Making Dataframe, prepare for model

### Major file: `SiteAdj`, the `deltaSOCAdj` (t/ha, not crop specific) is the final delta SOC

```{r Data preparation}

SiteData <- SiteAdj %>%
  select(Year = year, SiteName, Province = Province_ID, Ecodistrict = ECODISTRIC, Area = Area_per_site_Ha, deltaSOC = deltaSOCAdj )

#join the weather data
SiteData <- SiteData %>%
  inner_join(Weather, by = c("Ecodistrict" = "Ecodistrict" )) %>%
  inner_join(EcoP, by = c("Ecodistrict" =  "ECODISTRIC", "Province" = "PROVINCE","SiteName" = "sitename"))
  

#join the Cinput
SiteData <- SiteData %>%
  inner_join(TotalCinput, by = c( "Ecodistrict" = "ECODISTRIC", "Province" = "Province_ID", "Year" = "YEAR"))


#NA omit
SiteData <- na.omit(SiteData)

#Put dummy variables for soil texture and Eastern and Western Canada
#We believe crops in Western and Eastern Canada are quite different and have different C/N that could influence delta SOC
SiteData <- SiteData %>%
mutate(TEXT = case_when(
TEXCAT == "coarse" ~ 1,
TEXCAT == "medium" ~ 2,
TEXCAT == "fine" ~ 3,
TRUE ~ 99)) %>%
mutate(Region = case_when(
Province %in% c("BC","AB","SK","MB") ~ 1,
TRUE ~ 2))
```

## Build model 

### Linear model, Stepwise

```{r stepwise model}
full_model <- lm(deltaSOC ~ .-Year -SiteName -Province - Ecodistrict - Area, data = SiteData)
stepwise_model <- step(full_model, direction = "backward")
summary(stepwise_model)
cat("R² value: ", summary(stepwise_model)$r.squared, "\n")


model <- lm(deltaSOC ~ MAP + CARB30 + Cinput + TEXT + Region , data = SiteData)
stepwise_model <- step(model, direction = "backward")
summary(stepwise_model)

```


### Generalized additive model

```{r Generalized additive model}
# Fit a GAM model
model_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(TEXT, k=5) + s(CARB30) + s(Region, k = 5), data = SiteData)
model_gam_grow <- gam(deltaSOC ~ s(MAP) + s(MAT_grow) + s(MPE) + s(Cinput) + s(TCLAY) + s(TSAND) + s(CARB30), data = SiteData)

# Summary of the model
summary(model_gam) #R2 = 27.4
summary(model_gam_grow) #R2 = 27.7

```


```{r model comparision}
# AIC comparison
AIC(model, model_gam) #barely different
AIC(model_gam,model_gam_grow)#barely different
```


## Group data and redo the regression models

In this part, we will group the data into three soil texture groups (Fine, Medium, Coarse),
or C input (High, >64450 and low <= 64450), 
or MAP (High >1000. low <= 1000 )


### Soil textured group

```{r Grouped model- Soil texture}
#1. Soil texture group
SiteData_Fine <- SiteData %>%
  filter(TEXCAT == "fine")
SiteData_Medium <- SiteData %>%
  filter(TEXCAT == "medium")
SiteData_Coarse <- SiteData %>%
  filter(TEXCAT == "coarse")


#Fit the regression models
model_Fine <- lm(deltaSOC ~ MAP + MAT + CARB30 + Cinput + Region , data = SiteData_Fine)
stepwise_Fine <- step(model_Fine, direction = "backward")
summary(stepwise_Fine) # R2 = 0.140

model_Med <- lm(deltaSOC ~ MAP + MAT + CARB30 + Cinput + Region , data = SiteData_Medium)
stepwise_Med <- step(model_Med, direction = "backward")
summary(stepwise_Med) #R2 = 0.184

model_Coarse <- lm(deltaSOC ~ MAP + MAT + CARB30 + Cinput + Region , data = SiteData_Coarse)
stepwise_Coarse <- step(model_Coarse, direction = "backward")
summary(stepwise_Coarse) #highest R2= 0.398


#GAM 
model_Fine_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(CARB30), data = SiteData_Fine)
summary(model_Fine_gam) #R2 = 0.183

model_Med_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(CARB30), data = SiteData_Medium)
summary(model_Med_gam) #R2 = 0.232

model_Coarse_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(CARB30), data = SiteData_Coarse)
summary(model_Coarse_gam) #R2 = 0.429

```


### C input group

The C input ranges were determined by Q1, Q2, Q3, and Q4


``` {r Grouped model - C input}
#2. C input group
SiteData_HighC <- SiteData %>%
  filter(Cinput >= 248710 )
SiteData_MHC <- SiteData %>% 
  filter(Cinput >= 64450 & Cinput < 248710 ) #Medium high
SiteData_MLC <- SiteData %>%
  filter(Cinput <= 64450 & Cinput > 5020 ) #Medium low
SiteData_LowC <- SiteData %>%
  filter(Cinput <= 5020 )

#Fit the regression models
model_HighC <- lm(deltaSOC ~ MAP + MAT + CARB30 + Cinput + Region , data = SiteData_HighC)
stepwise_HighC <- step(model_HighC, direction = "backward")
summary(stepwise_HighC) #R2 = 0.127

model_MHC <- lm(deltaSOC ~ MAP + MAT + CARB30 + Cinput + Region , data = SiteData_MHC)
stepwise_MHC <- step(model_MHC, direction = "backward")
summary(stepwise_MHC) #R2 = 0.149

model_MLC <- lm(deltaSOC ~ MAP + MAT + CARB30 + Cinput + Region , data = SiteData_MLC)
stepwise_MLC <- step(model_MLC, direction = "backward")
summary(stepwise_MLC) #R2 = 0.144

model_LowC <- lm(deltaSOC ~ MAP + MAT + CARB30 + Cinput + Region , data = SiteData_LowC)
stepwise_LowC <- step(model_LowC, direction = "backward")
summary(stepwise_LowC) # R2 = 0.299

#GAM 
model_HighC_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(CARB30), data = SiteData_HighC)
summary(model_HighC_gam) #R2 = 0.150

model_MHC_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(CARB30), data = SiteData_MHC)
summary(model_MHC_gam) #R2 = 0.189

model_MLC_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(CARB30), data = SiteData_MLC)
summary(model_MLC_gam) #R2 = 0.188

model_LowC_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(CARB30), data = SiteData_LowC)
summary(model_LowC_gam) #R2 = 0.347

```

### MAP group

The MAPs were determined arbitary


```{r Group model - MAP}
#3. MAP group
SiteData_HighP <- SiteData %>%
  filter(MAP > 1000 )
SiteData_MedP <- SiteData %>%
  filter(MAP > 500 & MAP <= 1000 )
SiteData_LowP <- SiteData %>%
  filter(MAP <= 500 )

#Fit the regression models
model_HighP <- lm(deltaSOC ~ MAP + MAT + CARB30 + Cinput + Region , data = SiteData_HighP)
stepwise_HighP <- step(model_HighP, direction = "backward")
summary(stepwise_HighP) #R2 0.304

model_MedP <- lm(deltaSOC ~ MAP + MAT + CARB30 + Cinput + Region , data = SiteData_MedP)
stepwise_MedP <- step(model_MedP, direction = "backward")
summary(stepwise_MedP) #R2 0.243

model_LowP <- lm(deltaSOC ~ MAP + MAT + CARB30 + Cinput + Region , data = SiteData_LowP)
stepwise_LowP <- step(model_LowP, direction = "backward")
summary(stepwise_LowP) #R2 = 0.114


#GAM 
model_HighP_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(CARB30), data = SiteData_HighP)
summary(model_HighP_gam) #R2 = 0.365

model_MedP_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(CARB30), data = SiteData_MedP)
summary(model_MedP_gam) #R2 = 0.291

model_LowP_gam <- gam(deltaSOC ~ s(MAP) + s(MAT) + s(MPE) + s(Cinput) + s(CARB30), data = SiteData_LowP)
summary(model_LowP_gam) #R2 = 0.151

```


## Make a Lookup table

The look-up table include MAP, MAT, Soil Texture Class and C input
Based on the regression results, I believe =


```{r Lookup table }
# Ensure TEXTCAT is a factor
SiteData$TEXCAT <- as.factor(SiteData$TEXCAT)


# Bin MAP
SiteData$MAP_bin <- cut(SiteData$MAP, breaks = 4, labels = FALSE)
# Bin MAT
SiteData$MAT_bin <- cut(SiteData$MAT, breaks = 2, labels = FALSE)
# Bin Cinput
SiteData$Cinput_bin <- cut(SiteData$Cinput, breaks = 4, labels = FALSE)


# Create lookup table
lookup_table <- SiteData %>%
  group_by(MAP_bin, MAT_bin, Cinput_bin, TEXCAT) %>%
  summarize(mean_deltaSOC = mean(deltaSOC, na.rm = TRUE),
            median_deltaSOC = median(deltaSOC, na.rm = TRUE),
            sd_deltaSOC = sd(deltaSOC, na.rm = TRUE),
            n = n()) %>%
  ungroup()

```

